\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{hyperref}

\title{Post-typhoon OWT -- Architecture and Code Explanation (Simple Version)}
\author{}
\date{}

\begin{document}
\maketitle

\section{Purpose}
This memo explains a simplified implementation of the architecture required in pages 29--37 of the
PostTyphoon OWT document. The design preserves the mandatory 21-slot self-calibration module and
connects it to a transformer-style predictor of $f_0^{\mathrm{OWT}}$.

\section{System overview}
\textbf{Pipeline:}
\begin{itemize}
  \item Inputs: load history (time series) and soil/pile parameters.
  \item Slot-attention module: outputs $H_{ij}^0$ and 20 plastic drops $H_{ij}^n$ with gates $w_n(t)$.
  \item Stiffness evolution: $H_{ij}(t) = H_{ij}^0 - \sum_{n=1}^{20} w_n(t)\,H_{ij}^n$.
  \item Transformer encoder: consumes load tokens and stiffness-informed representation.
  \item Output: predicted $f_0^{\mathrm{OWT}}(t)$ sequence.
\end{itemize}

\section{Mandatory 21-slot module}
The 21 slots follow the physics-based macroelement template:
\begin{itemize}
  \item Slot 1 produces the initial 3DOF stiffness matrix $H_{ij}^0$.
  \item Slots 2--21 each produce a plastic drop matrix $H_{ij}^n$ and a gate parameterized by $w_n(t)$.
  \item Gates are sigmoid activations that enable smooth, monotone degradation.
\end{itemize}

\section{Simplified transformer}
The simplified model uses:
\begin{itemize}
  \item Tokenization of the load sequence with positional embeddings.
  \item Slot-attention to extract stiffness matrices.
  \item A transformer encoder block to capture temporal dependencies.
  \item A linear head to map embeddings to the predicted $f_0^{\mathrm{OWT}}(t)$.
\end{itemize}

\section{Physics consistency (simple)}
We enforce a basic physics consistency by comparing the predicted $f_0^{\mathrm{OWT}}$ to a proxy
frequency computed from the stiffness trace:
\begin{equation}
  f_0(t) \propto \sqrt{\mathrm{trace}(H_{ij}(t))}
\end{equation}
This keeps the prediction aligned with stiffness degradation trends.

\section{Code map}
The simplified implementation is provided in:
\begin{itemize}
  \item \texttt{meeting 3.2.1/simple\_slot\_transformer\_demo.py}
\end{itemize}
Key elements:
\begin{itemize}
  \item \textbf{SlotAttention}: cross-attention + GRU updates, yielding 21 slots.
  \item \textbf{SimpleSlotTransformer}: builds $H_{ij}(t)$ and predicts $f_0^{\mathrm{OWT}}(t)$.
  \item \textbf{demo()}: runs a forward pass and prints output shapes.
\end{itemize}

\section{How to run}
\begin{verbatim}
C:\Users\youss\Downloads\PFE\.venv\Scripts\python.exe "meeting 3.2.1\simple_slot_transformer_demo.py"
\end{verbatim}
Expected output:
\begin{verbatim}
ok torch.Size([8, 50]) torch.Size([8, 50])
\end{verbatim}

\section{Limitations}
This is a simplified, data-free version. It preserves the 21-slot requirement and the
core transformer flow, but does not include full experimental datasets or complete
validation protocols.

\end{document}
