\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xcolor}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  numbers=left,
  numberstyle=\tiny,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  stringstyle=\color{teal},
  showstringspaces=false
}

\title{Post-typhoon OWT -- Architecture and Code (Full + Simple)}
\author{}
\date{}

\begin{document}
\maketitle

\section{Architecture summary (pages 29--37)}
\textbf{Goal.} Predict typhoon-induced shifts of $f_0^{\mathrm{OWT}}$ using a physics-informed,
self-calibrating transformer.

\textbf{Key requirements:}
\begin{itemize}
  \item Scaled experiments with Froude similitude and realistic soil reproduction.
  \item A 21-slot self-calibration module (mandatory).
  \item Transformer encoder-decoder with masked self-attention and cross-attention.
  \item Physics losses enforcing LoT1 and LoT2 (frequency consistency and monotone
        non-negative stiffness degradation).
  \item Validation on separate data splits and attention-based XAI (heatmaps + roll-out).
\end{itemize}

\section{Full implementation (meeting 3.2)}
This version includes the full architecture, training pipeline, and XAI outputs.

\subsection{Model (slot-attention + encoder-decoder)}
\lstinputlisting[language=Python]{../meeting\ 3.2/model.py}

\subsection{Data simulation (FEM, analytical, experimental, field)}
\lstinputlisting[language=Python]{../meeting\ 3.2/data.py}

\subsection{Physics losses}
\lstinputlisting[language=Python]{../meeting\ 3.2/losses.py}

\subsection{Training and validation}
\lstinputlisting[language=Python]{../meeting\ 3.2/train.py}

\section{Simple implementation (meeting 3.2.1)}
This version keeps the mandatory 21-slot architecture and a simplified transformer.

\subsection{Simple 21-slot demo}
\lstinputlisting[language=Python]{./simple_slot_transformer_demo.py}

\section{How to run}
\begin{verbatim}
C:\Users\youss\Downloads\PFE\.venv\Scripts\python.exe "meeting 3.2\post_typhoon_slot_transformer.py"
C:\Users\youss\Downloads\PFE\.venv\Scripts\python.exe "meeting 3.2.1\simple_slot_transformer_demo.py"
\end{verbatim}

\end{document}
